---
title: "Applied Examples: Quarterly vs Monthly Labor Market Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applied Examples: Quarterly vs Monthly Labor Market Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5
)
```

## Introduction

Quarterly labor market data from PNADC represents **moving averages** of monthly
phenomena. This averaging smooths out short-term fluctuations, obscuring the
precise timing of economic turning points and the true magnitude of shocks.

This vignette demonstrates how `mensalizePNADC` reveals dynamics that quarterly
data hides. We focus on specific periods and comparisons where the difference
is most striking.

```{r load-data, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(scales)

# Load pre-computed series (included with package)
data("vignette_series", package = "mensalizePNADC")

# Prepare data
quarterly_total  <- as.data.table(vignette_quarterly_total)
quarterly_gender <- as.data.table(vignette_quarterly_gender)
quarterly_race   <- as.data.table(vignette_quarterly_race)
monthly_total    <- as.data.table(vignette_monthly_total)
monthly_gender   <- as.data.table(vignette_monthly_gender)
monthly_race     <- as.data.table(vignette_monthly_race)

cat("Period:", vignette_metadata$period_start, "to", vignette_metadata$period_end, "\n")
cat("Determination rate:", sprintf("%.1f%%", vignette_metadata$determination_rate * 100), "\n")
```

---

## 1. COVID-19: Revealing the True Shock

The COVID-19 pandemic caused the sharpest labor market shock in Brazilian history.
But quarterly data **understates the peak** and **delays the timing**.

```{r covid-unemployment, fig.height=5}
# Focus on COVID period
covid_q <- quarterly_total[period >= "2019-10-01" & period <= "2022-01-01"]
covid_m <- monthly_total[period >= "2019-10-01" & period <= "2022-01-01"]

# Find peaks
peak_quarterly <- covid_q[unemployment_rate == max(unemployment_rate)]
peak_monthly <- covid_m[unemployment_rate == max(unemployment_rate)]

ggplot() +
  # Quarterly as step function (shows moving average nature)
  geom_step(data = covid_q, aes(x = period, y = unemployment_rate),
            color = "steelblue", linewidth = 1.2, direction = "mid") +
  geom_point(data = covid_q, aes(x = period, y = unemployment_rate),
             color = "steelblue", size = 3) +

  # Monthly as line
  geom_line(data = covid_m, aes(x = period, y = unemployment_rate),
            color = "darkred", linewidth = 0.9) +
  geom_point(data = covid_m, aes(x = period, y = unemployment_rate),
             color = "darkred", size = 1.5) +

  # Annotate peaks
  annotate("segment", x = peak_monthly$period, xend = peak_monthly$period,
           y = peak_monthly$unemployment_rate + 0.005,
           yend = peak_monthly$unemployment_rate + 0.02,
           arrow = arrow(length = unit(0.2, "cm")), color = "darkred") +
  annotate("text", x = peak_monthly$period, y = peak_monthly$unemployment_rate + 0.022,
           label = paste0("Monthly peak: ", format(peak_monthly$period, "%b %Y"), "\n",
                         sprintf("%.1f%%", peak_monthly$unemployment_rate * 100)),
           size = 3.2, color = "darkred", hjust = 0.5) +

  annotate("segment", x = peak_quarterly$period + 30, xend = peak_quarterly$period + 30,
           y = peak_quarterly$unemployment_rate - 0.005,
           yend = peak_quarterly$unemployment_rate - 0.02,
           arrow = arrow(length = unit(0.2, "cm")), color = "steelblue") +
  annotate("text", x = peak_quarterly$period + 30, y = peak_quarterly$unemployment_rate - 0.025,
           label = paste0("Quarterly peak: ", format(peak_quarterly$period, "%b %Y"), "\n",
                         sprintf("%.1f%%", peak_quarterly$unemployment_rate * 100)),
           size = 3.2, color = "steelblue", hjust = 0.5) +

  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     limits = c(0.10, 0.17)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +

  labs(
    title = "COVID-19 Unemployment Spike: Monthly vs Quarterly",
    subtitle = "Monthly data (red) shows the true peak; quarterly (blue) averages it away",
    x = NULL, y = "Unemployment Rate",
    caption = "Source: PNADC/IBGE. Quarterly shown as step function to emphasize moving-average nature."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key findings:**

- Monthly data shows unemployment peaked at **`r sprintf("%.1f%%", peak_monthly$unemployment_rate * 100)`** in `r format(peak_monthly$period, "%B %Y")`
- Quarterly data shows a lower peak of **`r sprintf("%.1f%%", peak_quarterly$unemployment_rate * 100)`** - understating the shock
- The quarterly "peak" appears later than the true monthly peak

---

## 2. The 2014-2017 Recession: Tracking the Deterioration

Brazil's 2014-2017 recession saw unemployment rise from 6% to 14%. Monthly data
reveals the **month-by-month deterioration** that quarterly data smooths.

```{r recession-detail, fig.height=5.5}
# Focus on recession build-up
recession_q <- quarterly_total[period >= "2014-06-01" & period <= "2017-06-01"]
recession_m <- monthly_total[period >= "2014-06-01" & period <= "2017-06-01"]

# Find the peak
peak_m <- recession_m[unemployment_rate == max(unemployment_rate)]

# Calculate month-to-month changes for monthly
setorder(recession_m, period)
recession_m[, change := unemployment_rate - shift(unemployment_rate)]
big_jumps <- recession_m[!is.na(change) & change > 0.01]

ggplot() +
  # Quarterly
  geom_step(data = recession_q, aes(x = period, y = unemployment_rate),
            color = "steelblue", linewidth = 1, direction = "mid", alpha = 0.7) +
  geom_point(data = recession_q, aes(x = period, y = unemployment_rate),
             color = "steelblue", size = 2.5) +

  # Monthly
  geom_line(data = recession_m, aes(x = period, y = unemployment_rate),
            color = "darkred", linewidth = 0.8) +
  geom_point(data = recession_m, aes(x = period, y = unemployment_rate),
             color = "darkred", size = 1.2) +

  # Highlight big monthly jumps
  geom_point(data = big_jumps, aes(x = period, y = unemployment_rate),
             color = "darkred", size = 3, shape = 21, fill = "yellow", stroke = 1.5) +

  # Mark the peak
  annotate("point", x = peak_m$period, y = peak_m$unemployment_rate,
           color = "darkred", size = 4, shape = 18) +
  annotate("text", x = peak_m$period + 45, y = peak_m$unemployment_rate,
           label = paste0("Peak: ", format(peak_m$period, "%b %Y"), "\n",
                         sprintf("%.1f%%", peak_m$unemployment_rate * 100)),
           size = 3, hjust = 0, color = "darkred") +

  scale_y_continuous(labels = percent_format(accuracy = 1),
                     breaks = seq(0.06, 0.14, 0.02)) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +

  labs(
    title = "2014-2017 Recession: Month-by-Month Unemployment Rise",
    subtitle = "Yellow circles: months with >1 p.p. unemployment increase. Monthly data (red) vs quarterly (blue).",
    x = NULL, y = "Unemployment Rate",
    caption = "Source: PNADC/IBGE. Monthly data reveals sudden jumps hidden in quarterly averages."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key findings:**

- Monthly data reveals **`r nrow(big_jumps)` months** with unemployment jumps exceeding 1 percentage point
- These sudden deteriorations are invisible in quarterly data
- The precise peak occurred in **`r format(peak_m$period, "%B %Y")`** at `r sprintf("%.1f%%", peak_m$unemployment_rate * 100)`

---

## 3. Gender Gap During COVID: Women Left the Labor Force

COVID-19 disproportionately affected women's labor force participation due to
school closures and care responsibilities. Monthly data captures the **exact
timing and magnitude** of this divergence.

```{r gender-gap-covid, fig.height=5.5}
# Prepare gender data for COVID period
covid_gender_m <- monthly_gender[period >= "2019-10-01" & period <= "2021-09-01"]
covid_gender_q <- quarterly_gender[period >= "2019-10-01" & period <= "2021-09-01"]

# Calculate gender gap
gap_monthly <- dcast(covid_gender_m, period ~ sexo, value.var = "participation_rate")
gap_monthly[, gap := Men - Women]
setorder(gap_monthly, period)

# Find max gap
max_gap <- gap_monthly[gap == max(gap)]

# Pre-COVID baseline
baseline <- gap_monthly[period == "2020-02-15"]

ggplot() +
  # Men - monthly
  geom_line(data = covid_gender_m[sexo == "Men"],
            aes(x = period, y = participation_rate),
            color = "steelblue", linewidth = 1) +
  # Women - monthly
  geom_line(data = covid_gender_m[sexo == "Women"],
            aes(x = period, y = participation_rate),
            color = "coral", linewidth = 1) +

  # Shade the gap at maximum point
  annotate("rect",
           xmin = max_gap$period - 20, xmax = max_gap$period + 20,
           ymin = max_gap$Women, ymax = max_gap$Men,
           fill = "purple", alpha = 0.2) +
  annotate("segment",
           x = max_gap$period, xend = max_gap$period,
           y = max_gap$Women, yend = max_gap$Men,
           color = "purple", linewidth = 1.5) +
  annotate("text", x = max_gap$period + 25, y = (max_gap$Men + max_gap$Women) / 2,
           label = paste0("Max gap: ", sprintf("%.1f p.p.", (max_gap$Men - max_gap$Women) * 100)),
           size = 3, hjust = 0, color = "purple") +

  # Labels
  annotate("text", x = as.Date("2021-06-01"), y = 0.715, label = "Men",
           color = "steelblue", fontface = "bold", size = 4) +
  annotate("text", x = as.Date("2021-06-01"), y = 0.505, label = "Women",
           color = "coral", fontface = "bold", size = 4) +

  # Baseline reference
  geom_hline(yintercept = baseline$Women, linetype = "dashed", color = "coral", alpha = 0.5) +
  annotate("text", x = as.Date("2019-11-01"), y = baseline$Women + 0.01,
           label = "Pre-COVID\nwomen's rate", size = 2.5, color = "coral", hjust = 0) +

  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0.45, 0.75)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +

  labs(
    title = "COVID-19 Gender Impact: Women's Participation Collapsed",
    subtitle = paste0("Gender gap widened from ", sprintf("%.1f", baseline$gap * 100),
                     " to ", sprintf("%.1f", max_gap$gap * 100),
                     " p.p. - monthly data shows exact timing"),
    x = NULL, y = "Labor Force Participation Rate",
    caption = "Source: PNADC/IBGE. Monthly data from mensalizePNADC."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key findings:**

- Pre-COVID gender gap: **`r sprintf("%.1f", baseline$gap * 100)`** percentage points
- Maximum gap during COVID: **`r sprintf("%.1f", max_gap$gap * 100)`** percentage points in `r format(max_gap$period, "%B %Y")`
- Women's participation dropped from `r sprintf("%.1f%%", baseline$Women * 100)` to `r sprintf("%.1f%%", gap_monthly[Women == min(Women), Women] * 100)`
- Monthly data pinpoints when the divergence began and when recovery started

---

## 4. Racial Inequality: Unemployment Gap Widened During COVID

The unemployment gap between Black and White workers **widened significantly**
during the pandemic. Monthly data reveals when this happened.

```{r racial-gap, fig.height=5.5}
# Prepare race data - focus on unemployment rate gap
race_unemp_m <- monthly_race[period >= "2019-01-01" & period <= "2022-01-01"]
race_unemp_q <- quarterly_race[period >= "2019-01-01" & period <= "2022-01-01"]

# Calculate gap (Black - White)
gap_race_m <- dcast(race_unemp_m, period ~ raca, value.var = "unemployment_rate")
gap_race_m[, gap := Black - White]
setorder(gap_race_m, period)

# Find max gap
max_race_gap <- gap_race_m[gap == max(gap, na.rm = TRUE)]

# Pre-COVID
baseline_race <- gap_race_m[period == "2020-02-15"]

ggplot() +
  # White unemployment
  geom_line(data = race_unemp_m[raca == "White"],
            aes(x = period, y = unemployment_rate),
            color = "#2166AC", linewidth = 1) +
  # Black unemployment
  geom_line(data = race_unemp_m[raca == "Black"],
            aes(x = period, y = unemployment_rate),
            color = "#B2182B", linewidth = 1) +
  # Brown unemployment
  geom_line(data = race_unemp_m[raca == "Brown"],
            aes(x = period, y = unemployment_rate),
            color = "#DFC27D", linewidth = 1) +

  # Shade max gap period
  annotate("rect",
           xmin = max_race_gap$period - 20, xmax = max_race_gap$period + 20,
           ymin = max_race_gap$White, ymax = max_race_gap$Black,
           fill = "gray50", alpha = 0.2) +

  # COVID start marker
  geom_vline(xintercept = as.Date("2020-03-15"), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = as.Date("2020-03-15"), y = 0.21,
           label = "COVID\nstart", size = 2.5, hjust = -0.1) +

  # Labels
  annotate("text", x = as.Date("2021-10-01"), y = 0.175, label = "Black",
           color = "#B2182B", fontface = "bold", size = 3.5) +
  annotate("text", x = as.Date("2021-10-01"), y = 0.135, label = "Brown",
           color = "#DFC27D", fontface = "bold", size = 3.5) +
  annotate("text", x = as.Date("2021-10-01"), y = 0.095, label = "White",
           color = "#2166AC", fontface = "bold", size = 3.5) +

  # Gap annotation
  annotate("text", x = max_race_gap$period + 60, y = (max_race_gap$Black + max_race_gap$White) / 2,
           label = paste0("Max gap: ", sprintf("%.1f p.p.", max_race_gap$gap * 100), "\n",
                         format(max_race_gap$period, "%b %Y")),
           size = 3, hjust = 0) +

  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0.08, 0.22)) +
  scale_x_date(date_breaks = "4 months", date_labels = "%b\n%Y") +

  labs(
    title = "Racial Unemployment Gap: COVID Amplified Inequality",
    subtitle = paste0("Black-White gap widened from ", sprintf("%.1f", baseline_race$gap * 100),
                     " to ", sprintf("%.1f", max_race_gap$gap * 100), " p.p."),
    x = NULL, y = "Unemployment Rate",
    caption = "Source: PNADC/IBGE. Monthly data from mensalizePNADC."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key findings:**

- Pre-COVID Black-White gap: **`r sprintf("%.1f", baseline_race$gap * 100)`** percentage points
- Maximum gap: **`r sprintf("%.1f", max_race_gap$gap * 100)`** percentage points in `r format(max_race_gap$period, "%B %Y")`
- Black unemployment peaked at nearly **20%** while White unemployment peaked at **12%**
- Monthly data shows the gap widened immediately after COVID started and remained elevated

---

## 5. Formalization Rate: Detecting the "COVID Paradox"

During COVID, the formalization rate **increased** despite the economic collapse.
Why? Informal workers lost jobs faster, changing the composition of employment.

```{r formalization-covid, fig.height=5}
# Focus on COVID period formalization
covid_form_m <- monthly_total[period >= "2019-10-01" & period <= "2021-06-01",
                               .(period, formalization_rate)]
covid_form_q <- quarterly_total[period >= "2019-10-01" & period <= "2021-06-01",
                                 .(period, formalization_rate)]

# Find the spike
max_formal <- covid_form_m[formalization_rate == max(formalization_rate)]
pre_covid <- covid_form_m[period == "2020-02-15"]

ggplot() +
  # Quarterly
  geom_step(data = covid_form_q, aes(x = period, y = formalization_rate),
            color = "steelblue", linewidth = 1, direction = "mid") +
  geom_point(data = covid_form_q, aes(x = period, y = formalization_rate),
             color = "steelblue", size = 2.5) +

  # Monthly
  geom_line(data = covid_form_m, aes(x = period, y = formalization_rate),
            color = "darkgreen", linewidth = 0.9) +
  geom_point(data = covid_form_m, aes(x = period, y = formalization_rate),
             color = "darkgreen", size = 1.5) +

  # Highlight the spike
  annotate("point", x = max_formal$period, y = max_formal$formalization_rate,
           color = "darkgreen", size = 4, shape = 18) +
  annotate("segment", x = max_formal$period, xend = max_formal$period,
           y = max_formal$formalization_rate + 0.005,
           yend = max_formal$formalization_rate + 0.025,
           arrow = arrow(length = unit(0.2, "cm")), color = "darkgreen") +
  annotate("text", x = max_formal$period, y = max_formal$formalization_rate + 0.03,
           label = paste0("COVID spike:\n", sprintf("%.1f%%", max_formal$formalization_rate * 100), "\n",
                         "(informal jobs lost)"),
           size = 3, hjust = 0.5, color = "darkgreen") +

  # Pre-COVID reference
  geom_hline(yintercept = pre_covid$formalization_rate, linetype = "dashed", alpha = 0.5) +

  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0.56, 0.66)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +

  labs(
    title = "The COVID Formalization Paradox",
    subtitle = "Formalization rate ROSE because informal workers lost jobs disproportionately",
    x = NULL, y = "Formalization Rate",
    caption = "Source: PNADC/IBGE. Green = monthly, blue = quarterly."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key findings:**

- Pre-COVID formalization: **`r sprintf("%.1f%%", pre_covid$formalization_rate * 100)`**
- COVID peak: **`r sprintf("%.1f%%", max_formal$formalization_rate * 100)`** in `r format(max_formal$period, "%B %Y")`
- This `r sprintf("%.1f", (max_formal$formalization_rate - pre_covid$formalization_rate) * 100)` p.p. jump happened in a single month
- Monthly data captures this compositional shift that quarterly data smooths

---

## 6. Minimum Wage Adjustments: Habitual vs Effective Income

One of the most compelling demonstrations of mensalization's value is capturing **minimum wage adjustments**. In PNADC:

- **Habitual income (VD4016)**: What the worker usually earns - refers to the **current** reference month
- **Effective income (VD4017)**: What the worker actually received - refers to the **previous** month

When the minimum wage increases (typically in January), **habitual income captures the change immediately**, while **effective income shows it with a one-month lag**. This timing difference is invisible in quarterly data but becomes clear with monthly mensalization.

```{r mw-analysis-setup, message=FALSE}
# Load minimum wage data
monthly_mw <- as.data.table(vignette_monthly_mw)
mw_history <- as.data.table(vignette_mw_history)

# Create a long format for plotting
mw_long <- melt(monthly_mw,
                id.vars = c("period", "ref_month_yyyymm", "mw_value", "is_adjustment_month"),
                measure.vars = c("pct_at_mw_habitual", "pct_at_mw_effective"),
                variable.name = "income_type",
                value.name = "pct_at_mw")

mw_long[, income_label := fifelse(income_type == "pct_at_mw_habitual",
                                   "Habitual Income\n(current month)",
                                   "Effective Income\n(previous month)")]
```

```{r mw-timeseries, fig.height=6, fig.width=10}
# Get MW adjustment months for vertical lines
adj_months <- monthly_mw[is_adjustment_month == TRUE, period]

ggplot(mw_long, aes(x = period, y = pct_at_mw, color = income_label)) +
  geom_line(linewidth = 0.8) +

  # Mark MW adjustment months
  geom_vline(xintercept = adj_months, linetype = "dashed", alpha = 0.4, color = "gray40") +

  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_color_manual(values = c("Habitual Income\n(current month)" = "steelblue",
                                "Effective Income\n(previous month)" = "coral")) +

  labs(
    title = "Share of Formal Workers Earning Minimum Wage (+/- 5%)",
    subtitle = "Dashed lines mark MW adjustment months. Habitual income shows MW changes immediately; effective income shows 1-month lag.",
    x = NULL, y = "Share of Workers",
    color = "Income Measure",
    caption = "Source: PNADC/IBGE. Formal private sector employees only. MW values from IPEADATA."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Focus: January 2020 MW Adjustment (R$ 998 to R$ 1,039)

The 2020 adjustment provides a clear example. In January 2020, the MW increased from R$ 998 to R$ 1,039. Let's zoom in:

```{r mw-2020-zoom, fig.height=5}
# Focus on late 2019 to mid 2020
mw_2020 <- mw_long[period >= "2019-09-01" & period <= "2020-06-01"]

# Get the specific adjustment month
adj_2020 <- as.Date("2020-01-15")

ggplot(mw_2020, aes(x = period, y = pct_at_mw, color = income_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +

  # Mark the adjustment
  geom_vline(xintercept = adj_2020, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = adj_2020, y = max(mw_2020$pct_at_mw) * 1.02,
           label = "MW: R$ 998 -> R$ 1,039", size = 3, hjust = -0.05) +

  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%Y") +
  scale_color_manual(values = c("Habitual Income\n(current month)" = "steelblue",
                                "Effective Income\n(previous month)" = "coral")) +

  labs(
    title = "MW Adjustment: January 2020",
    subtitle = "Habitual income shows spike in January; effective income shows spike in February",
    x = NULL, y = "Share at MW (+/- 5%)",
    color = NULL,
    caption = "Habitual income refers to current month; effective income refers to previous month."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

**Key insight**: In January 2020:
- **Habitual income** immediately shows workers at the new MW (R$ 1,039)
- **Effective income** still shows workers at the old MW (R$ 998) because it refers to December 2019

This one-month lag pattern repeats at every MW adjustment and demonstrates:

1. **The importance of monthly data**: This pattern is completely smoothed away in quarterly data
2. **Understanding income concepts**: Habitual vs effective income have different temporal references
3. **Policy evaluation**: When assessing minimum wage impacts, the correct timing matters

### MW Adjustment Events Summary

```{r mw-table}
# Calculate the jump in habitual vs effective income around each adjustment
setorder(monthly_mw, period)
monthly_mw[, hab_change := pct_at_mw_habitual - shift(pct_at_mw_habitual)]
monthly_mw[, eff_change := pct_at_mw_effective - shift(pct_at_mw_effective)]
monthly_mw[, hab_change_next := shift(pct_at_mw_habitual, type = "lead") - pct_at_mw_habitual]
monthly_mw[, eff_change_next := shift(pct_at_mw_effective, type = "lead") - pct_at_mw_effective]

# Get adjustment months with the changes
adj_summary <- monthly_mw[is_adjustment_month == TRUE & period >= "2013-01-01",
                           .(Month = format(period, "%b %Y"),
                             `MW Value (R$)` = mw_value,
                             `Habitual Jump (p.p.)` = sprintf("%+.1f", hab_change * 100),
                             `Effective Jump (p.p.)` = sprintf("%+.1f", eff_change * 100))]

knitr::kable(adj_summary,
             caption = "Minimum wage adjustments and immediate changes in share at MW",
             align = c("l", "r", "r", "r"))
```

---

## Summary: What Monthly Data Reveals

| Analysis | Quarterly Limitation | Monthly Insight |
|----------|---------------------|-----------------|
| **COVID unemployment** | Peak understated | True peak of `r sprintf("%.1f%%", peak_monthly$unemployment_rate * 100)` in `r format(peak_monthly$period, "%b %Y")` |
| **Recession build-up** | Gradual rise appears smooth | `r nrow(big_jumps)` months with >1 p.p. jumps |
| **Gender gap** | Gap widening timing unclear | Exact month of divergence (`r format(max_gap$period, "%b %Y")`) |
| **Racial inequality** | Gap changes smoothed | `r sprintf("%.1f", max_race_gap$gap * 100)` p.p. peak gap dated precisely |
| **Formalization paradox** | Spike diluted across quarter | Sharp compositional shift visible |
| **Minimum wage** | MW adjustment timing lost | 1-month lag between habitual and effective income captured |

Monthly data from `mensalizePNADC` enables:

- **Precise timing** of economic turning points
- **True magnitudes** of shocks (not moving averages)
- **Exact dating** of policy-relevant changes
- **Short-term dynamics** invisible in quarterly data

---

## Technical Notes

### Data Source

The series shown here were computed from **`r format(vignette_metadata$n_observations, big.mark = ",")`** PNADC observations covering `r vignette_metadata$period_start` to `r vignette_metadata$period_end`.

The mensalization algorithm achieved a **`r sprintf("%.1f%%", vignette_metadata$determination_rate * 100)`** determination rate, meaning nearly all observations could be assigned to specific months.

### Reproduce These Results

```r
library(mensalizePNADC)
library(data.table)

# Load your stacked PNADC data
pnadc <- fread("your_pnadc_stacked.csv")

# Apply mensalization with weight computation
result <- mensalizePNADC(pnadc, compute_weights = TRUE)

# Compute monthly unemployment rate
monthly_unemp <- result[!is.na(weight_monthly), .(
  unemployment_rate = sum((VD4001 == 1 & VD4002 == 2) * weight_monthly) /
                      sum((VD4001 == 1) * weight_monthly)
), by = ref_month_yyyymm]
```

See the [Getting Started vignette](getting-started.html) for the complete algorithm explanation.

---

## References

- IBGE. Pesquisa Nacional por Amostra de Domicilios Continua (PNADC).
  https://www.ibge.gov.br/estatisticas/sociais/trabalho/
- Hecksher, M. (2024). Mensalizacao da PNADC. Working paper.
