---
title: "Getting Started with mensalizePNADC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mensalizePNADC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `mensalizePNADC` package converts Brazil's quarterly PNADC (Pesquisa Nacional por Amostra de Domicilios Continua) survey data into monthly time series. It provides two main capabilities:

1. **Reference month identification**: Determines which month within each quarter each survey observation refers to
2. **Monthly weight computation**: Adjusts survey weights for monthly (instead of quarterly) estimates

### Performance Summary

Tested on 55 quarters (2012Q1 - 2025Q3) with 28,395,273 observations:

| Metric | Result |
|--------|--------|
| Overall determination rate | **95.83%** |
| Processing time (basic mode) | 6.3 minutes |
| Processing time (with weights) | 8.1 minutes |
| Best period (2013-2019) | 96-99% determination |

## Installation

```{r installation}
# Install from GitHub
devtools::install_github("antrologos/mensalizePNADC")
```

## Important: Use Stacked Data

The algorithm achieves **95.83% determination rate** when processing **stacked multi-quarter data**. If you process quarters individually, you'll only get ~65-75% determination.

| Processing Mode | Determination Rate |
|-----------------|-------------------|
| Per-quarter (single quarter) | 65-75% |
| Stacked (multi-quarter) | **95.83%** |

**Recommended**: Stack at least 2 years of quarterly data before calling `mensalizePNADC()`.

---

## Algorithm Overview: Visual Diagram

The reference month identification follows a 6-step pipeline. Here's the complete flow:

```
┌─────────────────────────────────────────────────────────────────────────────┐
│                    REFERENCE MONTH IDENTIFICATION PIPELINE                   │
└─────────────────────────────────────────────────────────────────────────────┘

INPUTS (per observation):
┌──────────────┐  ┌──────────────┐  ┌──────────────┐  ┌──────────────┐
│  Year (Ano)  │  │  Quarter     │  │  Birthday    │  │  Age         │
│              │  │  (Trimestre) │  │  (V2008/1/2) │  │  (V2009)     │
└──────────────┘  └──────────────┘  └──────────────┘  └──────────────┘
        │                │                  │                 │
        ▼                ▼                  ▼                 ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 1: Calculate Valid Interview Saturdays                                 │
│         (IBGE "Parada Técnica" Rules)                                       │
├─────────────────────────────────────────────────────────────────────────────┤
│  For each month, find first Saturday with ≥4 days in that month.            │
│  This defines when interviews for that month's reference week can occur.    │
│                                                                             │
│  OUTPUT: date_min = First Saturday of Month 1                               │
│          date_max = First Saturday of Month 3 + 21 days                     │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 2: Apply Birthday Constraints                                          │
├─────────────────────────────────────────────────────────────────────────────┤
│  Compare (Survey_Year - Birth_Year) with Reported_Age to determine if      │
│  interview was BEFORE or AFTER the person's birthday this year.            │
│                                                                             │
│  If AFTER birthday  → date_min = max(date_min, Sat_after_birthday)          │
│  If BEFORE birthday → date_max = min(date_max, Sat_before_birthday)         │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 3: Convert Date Ranges to Month Positions (1, 2, or 3)                 │
├─────────────────────────────────────────────────────────────────────────────┤
│  month_min_pos = which month does date_min fall in?                         │
│  month_max_pos = which month does date_max fall in?                         │
│                                                                             │
│  Special handling: days 1-3 of a month belong to previous month's week      │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 4: Aggregate to UPA-Panel Level                                        │
├─────────────────────────────────────────────────────────────────────────────┤
│  All people in same (UPA, V1014) are interviewed in the SAME MONTH.         │
│  Take intersection of all individual constraints:                           │
│                                                                             │
│    upa_month_min = max(all individual month_min_pos)                        │
│    upa_month_max = min(all individual month_max_pos)                        │
│                                                                             │
│  If upa_month_min = upa_month_max → DETERMINED!                             │
│  If upa_month_min < upa_month_max → Still ambiguous, try Step 5             │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 5: Cross-Quarter Aggregation (KEY INNOVATION)                          │
├─────────────────────────────────────────────────────────────────────────────┤
│  PNADC uses ROTATING PANEL: same (UPA, V1014) interviewed in SAME MONTH     │
│  POSITION across all quarterly visits (up to 5 consecutive quarters).       │
│                                                                             │
│  Combine constraints from ALL quarters → dramatically improves accuracy:    │
│                                                                             │
│    Per-quarter only:  ~65-75% determination                                 │
│    Cross-quarter:     ~95.83% determination                                 │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ STEP 6: Handle Exception Quarters                                           │
├─────────────────────────────────────────────────────────────────────────────┤
│  For quarters with non-standard timing (2016t3, 2016t4, 2017t2, 2022t3,     │
│  2023t2), apply relaxed rule: ≥3 days instead of ≥4 days.                   │
└─────────────────────────────────────────────────────────────────────────────┘
                                     │
                                     ▼
┌─────────────────────────────────────────────────────────────────────────────┐
│ OUTPUT                                                                       │
├─────────────────────────────────────────────────────────────────────────────┤
│  ref_month_in_quarter: 1, 2, 3, or NA (indeterminate)                       │
│  ref_month:            Full date (e.g., 2023-02-01)                         │
│  ref_month_yyyymm:     Integer format (e.g., 202302)                        │
└─────────────────────────────────────────────────────────────────────────────┘
```

---

## Basic Usage: Identifying Reference Months

The most common use case is identifying which month each observation refers to, then joining this information with your data.

### Step 1: Prepare your data

You need stacked quarterly PNADC data with at least these columns:

| Column | Description |
|--------|-------------|
| `Ano` | Survey year |
| `Trimestre` | Quarter (1-4) |
| `UPA` | Primary Sampling Unit |
| `V1014` | Panel identifier |
| `V1008` | Household identifier |
| `V2003` | Person identifier |
| `V2008` | Birth day (1-31, or 99 for unknown) |
| `V20081` | Birth month (1-12, or 99 for unknown) |
| `V20082` | Birth year (or 9999 for unknown) |
| `V2009` | Age |

```{r load-data}
library(mensalizePNADC)
library(data.table)

# Load your stacked quarterly PNADC data
pnadc <- fread("pnadc_stacked.csv",
  select = c("Ano", "Trimestre", "UPA", "V1008", "V1014", "V2003",
             "V2008", "V20081", "V20082", "V2009"))

# Check data dimensions
cat("Rows:", nrow(pnadc), "\n")
cat("Quarters:", uniqueN(pnadc[, .(Ano, Trimestre)]), "\n")
```

### Step 2: Get the crosswalk

```{r get-crosswalk}
crosswalk <- mensalizePNADC(pnadc)
# Step 1/1: Identifying reference months...
#   Determination rate: 95.8%

print(crosswalk)
# PNADC Reference Month Crosswalk
# -------------------------------
# Observations: 28,395,273
# Determination rate: 95.8%
# Date range: 201201 - 202509
#
# Join keys: Ano, Trimestre, UPA, V1008, V1014, V2003
# Output columns: ref_month, ref_month_in_quarter, ref_month_yyyymm
```

### Step 3: Join with original data

```{r join-data}
# Load an original quarterly file
library(haven)
original <- read_dta("PNADC_2023T1.dta")

# Join to add monthly information
monthly_data <- merge(original, crosswalk,
  by = c("Ano", "Trimestre", "UPA", "V1008", "V1014", "V2003"),
  all.x = TRUE)

# Now you have:
# - ref_month: Reference month as Date (e.g., 2023-01-01)
# - ref_month_in_quarter: Position in quarter (1, 2, 3, or NA)
# - ref_month_yyyymm: Integer YYYYMM format (e.g., 202301)
```

### Step 4: Use the monthly information

```{r use-monthly}
# Filter to a specific month
jan_2023 <- monthly_data[ref_month_yyyymm == 202301]

# Group by month
by_month <- monthly_data[, .(
  n_obs = .N,
  mean_age = mean(V2009, na.rm = TRUE)
), by = ref_month_yyyymm]

# Check determination rate by year
monthly_data[, .(
  total = .N,
  determined = sum(!is.na(ref_month_in_quarter)),
  rate = round(mean(!is.na(ref_month_in_quarter)) * 100, 1)
), by = Ano]
```

---

## How the Algorithm Works: Detailed Explanation

The reference month identification algorithm determines which of the three months in a quarter each survey observation belongs to. This section explains each step in detail with concrete examples.

### Background: PNADC Survey Structure

PNADC interviews households in **reference weeks** throughout each quarter. Each reference week:

- Ends on a **Saturday**
- Belongs to a specific **reference month** (the month containing most of the week)
- Follows IBGE's "Parada Técnica" (technical break) scheduling rules

The challenge: PNADC microdata only records the **quarter** (Trimestre), not the specific month. This package recovers the month using birthday constraints and panel structure.

---

### Step 1: Calculate Valid Interview Saturdays

**Goal**: Determine which Saturdays can serve as the end of a reference week for each month.

**IBGE Rule**: A reference week belongs to a month only if that Saturday has **at least 4 days** within the month. This ensures the majority of the 7-day reference week falls within that month.

**How we calculate the first valid Saturday**:

```
first_saturday_day = day of the first Saturday of the month
if first_saturday_day >= 4:
    use this Saturday (it has enough days in the month)
else:
    use the second Saturday (first_saturday_day + 7)
```

**Detailed Example for Q1 2023**:

```
JANUARY 2023:
  Sun  Mon  Tue  Wed  Thu  Fri  SAT
   1    2    3    4    5    6   [7]  ← First Saturday is day 7
                                     7 ≥ 4? YES → Valid! (7 days in January)

FEBRUARY 2023:
  Wed  Thu  Fri  SAT  ...
   1    2    3   [4]              ← First Saturday is day 4
                                  4 ≥ 4? YES → Valid! (4 days in February)

MARCH 2023:
  Wed  Thu  Fri  SAT  ...
   1    2    3   [4]              ← First Saturday is day 4
                                  4 ≥ 4? YES → Valid! (4 days in March)
```

**Counter-example showing when we skip to second Saturday**:

```
APRIL 2023:
  SAT  Sun  Mon  Tue  ...
  [1]   2    3    4              ← First Saturday is day 1
                                  1 ≥ 4? NO → Skip!

  Fri  SAT  Sun  ...
   7   [8]   9                   ← Second Saturday is day 8
                                  Use this one instead
```

**Why this matters**: The valid Saturday calculation defines the **possible interview date range** for the quarter:

- `date_min` = First valid Saturday of Month 1
- `date_max` = First valid Saturday of Month 3 + 21 days (allowing for 4 reference weeks)

For Q1 2023: date_min = Jan 7, date_max = Mar 4 + 21 = Mar 25.

---

### Step 2: Apply Birthday Constraints

**Goal**: Narrow the possible interview date window using each person's birthday and reported age.

**Key insight**: The reported age combined with birth year tells us whether the interview happened before or after the person's birthday this year.

**The calculation**:

```
visit_before_birthday = (Survey_Year - Birth_Year) - Reported_Age

If = 0: Interview was AFTER birthday (person already celebrated this year)
If = 1: Interview was BEFORE birthday (birthday hasn't happened yet)
```

**Detailed Example 1: Interview AFTER birthday**

```
Person: Born March 15, 1990
Survey Year: 2023
Reported Age: 33

Check: 2023 - 1990 = 33
       33 - 33 = 0  → Interview was AFTER March 15, 2023

Constraint: date_min = max(date_min, first_Saturday_on_or_after(March 15))
            date_min = max(Jan 7, March 18) = March 18

This person can only have been interviewed in MARCH (month 3).
The first two months are ruled out!
```

**Detailed Example 2: Interview BEFORE birthday**

```
Person: Born March 15, 1990
Survey Year: 2023
Reported Age: 32

Check: 2023 - 1990 = 33
       33 - 32 = 1  → Interview was BEFORE March 15, 2023

Constraint: date_max = min(date_max, Saturday_before(March 15))
            date_max = min(March 25, March 11) = March 11

This person can only have been interviewed in JANUARY or FEBRUARY.
March is ruled out!
```

**Why Saturdays?** Since reference weeks end on Saturdays, we constrain to the first Saturday after (or before) the birthday, not the birthday itself. This aligns with IBGE's scheduling.

**Unknown birthdays**: When V2008=99, V20081=99, or V20082=9999 (unknown), that person's birthday cannot constrain the date. However, they may still be determined through household aggregation (Step 4) or cross-quarter aggregation (Step 5).

---

### Step 3: Convert Date Ranges to Month Positions

**Goal**: Transform the date window [date_min, date_max] into month-in-quarter positions (1, 2, or 3).

**Basic conversion**:

```
month_min_pos = which month of the quarter does date_min fall in?
month_max_pos = which month of the quarter does date_max fall in?
```

**Boundary handling (days 1-3 of a month)**:

Interviews on days 1-3 of a month typically belong to a reference week that **started in the previous month**. The algorithm handles this:

```
For date_min: if day ≤ 3 AND not in first month of quarter:
              subtract 1 from month position

For date_max: if day ≤ 3:
              use the month of (date - 3 days) instead
```

**Example**:

```
date_min = March 2, 2023 (Q1)
  Day 2 ≤ 3 and March is not month 1 of Q1
  → month_min_pos = 3 - 1 = 2 (February)

date_max = March 25, 2023 (Q1)
  Day 25 > 3
  → month_max_pos = 3 (March)

Result: month_min_pos=2, month_max_pos=3 → still ambiguous (Feb or Mar)
```

**Determination check**:

```
If month_min_pos = month_max_pos → Reference month DETERMINED
If month_min_pos < month_max_pos → Still ambiguous, need more constraints
If month_min_pos > month_max_pos → Inconsistent (contradictory birthday data)
```

---

### Step 4: Aggregate to UPA-Panel Level

**Goal**: Combine constraints from all household members interviewed together.

**Key insight**: All people in the same **UPA + V1014 (panel)** are interviewed together in the **same reference month**. Their individual constraints with respect to the month must all be satisfied simultaneously.

**The aggregation**:

```
upa_month_min = MAX of all individual month_min_pos
upa_month_max = MIN of all individual month_max_pos
```

**Why this works**: Taking the maximum of minimums and minimum of maximums gives us the **intersection** of all individual constraint windows.

**Detailed Example**:

```
Household (UPA=123456, V1014=1) has 3 members:

Person A: month_min=1, month_max=2 (could be Jan or Feb)
Person B: month_min=1, month_max=3 (could be Jan, Feb, or Mar)
Person C: month_min=2, month_max=3 (could be Feb or Mar)

Aggregation:
  upa_month_min = max(1, 1, 2) = 2
  upa_month_max = min(2, 3, 3) = 2

Result: min=2, max=2 → Reference month is FEBRUARY!

Visual representation:
              Jan    Feb    Mar
Person A:     [======]
Person B:     [=============]
Person C:            [========]
Intersection:        [==]      ← Only February satisfies all constraints
```

**What if constraints are inconsistent?**

```
Person A: month_min=1, month_max=1 (must be Jan)
Person B: month_min=3, month_max=3 (must be Mar)

upa_month_min = max(1, 3) = 3
upa_month_max = min(1, 3) = 1

Result: min=3 > max=1 → INCONSISTENT!

This is rare and usually indicates data quality issues.
These observations remain indeterminate (ref_month_in_quarter = NA).
```

---

### Step 5: Cross-Quarter Aggregation (The Key Innovation)

**Goal**: Leverage PNADC's rotating panel design to combine constraints across quarters.

**PNADC Panel Structure**:

PNADC uses a rotating panel where each household (UPA + V1014) is interviewed for **5 consecutive quarters**. Crucially, the same household is always interviewed in the **same relative month position within the quarter** across all visits (always month 1, always month 2, or always month 3).

```
UPA=123456, V1014=1 interview schedule:
  2023-Q1: Month 1 of the quarter → January
  2023-Q2: Month 1 of the quarter → April
  2023-Q3: Month 1 of the quarter → July
  2023-Q4: Month 1 of the quarter → October
  2024-Q1: Month 1 of the quarter → January
           [Then rotates out]
```

**The aggregation logic**:

Since the relative month position is constant across quarters, constraints from **any** quarter apply to **all** quarters for that UPA-V1014:

```
For each (UPA, V1014) group across ALL quarters:
  upa_month_min = MAX of all month_min_pos from all quarters
  upa_month_max = MIN of all month_max_pos from all quarters
```

**Detailed Example**:

```
UPA=123456, V1014=1 appears in 5 quarters:

  2023-Q1: month_min=1, month_max=2 (ambiguous: Jan or Feb)
  2023-Q2: month_min=1, month_max=3 (ambiguous: Apr, May, or Jun)
  2023-Q3: month_min=2, month_max=2 (DETERMINED: Aug, i.e., month 2)
  2023-Q4: month_min=1, month_max=2 (ambiguous: Oct or Nov)
  2024-Q1: month_min=2, month_max=3 (ambiguous: Feb or Mar)

Cross-quarter aggregation:
  upa_month_min = max(1, 1, 2, 1, 2) = 2
  upa_month_max = min(2, 3, 2, 2, 3) = 2

Result: ALL 5 quarters get ref_month_in_quarter = 2

This means:
  2023-Q1 → February
  2023-Q2 → May
  2023-Q3 → August
  2023-Q4 → November
  2024-Q1 → February
```

**Why this is so powerful**:

A household that is ambiguous in Q1 (could be month 1 or 2) might have a birthday in Q3 that conclusively determines month 2. That determination propagates back to Q1!

```
┌─────────────────────────────────────────────────────────────────────┐
│ IMPACT ON DETERMINATION RATE:                                      │
│                                                                     │
│   Per-quarter processing (each quarter alone):    ~65-75%          │
│   Stacked multi-quarter (cross-quarter):          ~95.83%          │
│                                                                     │
│   Improvement: +25-30 percentage points!                            │
└─────────────────────────────────────────────────────────────────────┘
```

**This is why you should always stack multiple quarters before calling `mensalizePNADC()`!**

---

### Step 6: Handle Exception Quarters

**Goal**: Apply relaxed rules for quarters where IBGE used non-standard scheduling.

**Background**: In some quarters, IBGE's technical breaks crossed month boundaries in unusual ways. The standard rule (≥4 days) may produce impossible results for these quarters.

**Exception quarters**:

```{r exceptions}
get_exception_quarters()
# [1] "2016t3" "2016t4" "2017t2" "2022t3" "2023t2"
```

**Relaxed rule**: For observations still indeterminate after Step 5 in exception quarters, the algorithm:

1. Recalculates first valid Saturdays with **min_days = 3** (instead of 4)
2. Recalculates date ranges and applies birthday constraints
3. Recalculates month positions with adjusted boundary handling (day ≤ 2 instead of ≤ 3)
4. Re-aggregates to UPA-Panel level
5. If now determined, uses this result; otherwise remains NA

**Example**: In 2016t3, a Saturday that falls on day 3 of a month would be skipped under the standard rule (3 < 4) but accepted under the exception rule (3 ≥ 3).

---

### Summary: What Makes Observations Indeterminate?

After all 6 steps, approximately **4.2% of observations** remain indeterminate (ref_month_in_quarter = NA). Common reasons:

| Reason | Description |
|--------|-------------|
| All members have unknown birthdays | No birthday constraints available for the household |
| Single-quarter UPA-V1014 | Panel appears in only one quarter with insufficient constraints |
| Contradictory constraints | Birthday data is inconsistent (rare, indicates data quality issues) |
| Insufficient cross-quarter overlap | Not enough informative birthdays across the panel's visits |

**Handling indeterminate observations**:

- **Exclude**: Remove from monthly analysis
- **Quarterly fallback**: Use original quarterly weights for these observations
- **Proportional distribution**: Distribute 1/3 to each of the three months

---

## Computing Monthly Weights

For monthly aggregate estimates (e.g., monthly employment counts), you need monthly-appropriate survey weights.

### Required Additional Variables

| Variable | Description |
|----------|-------------|
| `V1028` | Original quarterly survey weight |
| `UF` | State code |
| `posest` | Post-stratification cell |
| `posest_sxi` | Post-stratification group |

### How Weight Calibration Works

When `compute_weights = TRUE`, the package:

1. **Fetches monthly population from SIDRA API** (table 6022)
   - SIDRA provides moving-quarter estimates
   - Package transforms to exact monthly values
   - Boundary months extrapolated via quadratic regression

2. **Applies hierarchical rake weighting** across nested cells:
   - `celula1`: Age groups (0-13, 14-29, 30-59, 60+)
   - `celula2`: Post-stratum group + age
   - `celula3`: State (UF) + celula2
   - `celula4`: Post-stratum (posest) + celula2

3. **Calibrates to monthly population totals**

```{r monthly-weights}
# Load full data with all required variables
pnadc_full <- fread("pnadc_stacked_full.csv")

# Run mensalization with weight computation
# (automatically fetches population from SIDRA API)
result <- mensalizePNADC(pnadc_full,
  compute_weights = TRUE,
  verbose = TRUE)
# Step 1/4: Identifying reference months...
#   Determination rate: 95.8%
# Step 2/4: Fetching monthly population from SIDRA...
# Step 3/4: Calibrating monthly weights (rake weighting)...
# Step 4/4: Smoothing monthly aggregates...

# Use weight_monthly for monthly estimates
monthly_pop <- result[, .(
  population = sum(weight_monthly, na.rm = TRUE)
), by = ref_month_yyyymm]
```

The `weight_monthly` output is calibrated to match IBGE's official monthly population estimates (~206 million average).

---

## Topic-Specific Calibration (Coming Soon)

A future version will include `calibrate_to_sidra()` for **theme-specific Bayesian calibration** to exactly match IBGE's published SIDRA series for specific indicators.

### How It Will Work

The base `weight_monthly` provides general-purpose monthly weights calibrated to population totals. However, different labor market indicators (unemployment rate, employment level, income) may require slightly different weight adjustments to exactly match IBGE's official published series.

The `calibrate_to_sidra()` function will:

1. **Fetch target series** from SIDRA (e.g., unemployment rate, employment levels)
2. **Apply Bayesian adjustment** to weights so aggregated estimates match official series
3. **Return theme-specific weights** (e.g., `weight_sidra_unemployment`)

**Planned themes:**

| Theme | What's Calibrated |
|-------|-------------------|
| `unemployment` | Unemployment rate matches IBGE exactly |
| `employment` | Employment levels match IBGE exactly |
| `labor_force` | Labor force participation matches IBGE |
| `income` | Income aggregates match IBGE |
| `custom` | User supplies their own target series |

---

## Understanding the Output

### Reference Month Variables

| Variable | Type | Description |
|----------|------|-------------|
| `ref_month` | Date | First day of reference month (e.g., "2023-01-01") |
| `ref_month_in_quarter` | Integer | Position: 1, 2, 3, or NA if indeterminate |
| `ref_month_yyyymm` | Integer | YYYYMM format (e.g., 202301) |
| `weight_monthly` | Numeric | Monthly weight (if `compute_weights = TRUE`) |

### Determination Rates by Period

| Period | Quarters | Determination Rate | Notes |
|--------|----------|-------------------|-------|
| 2012 | Q1-Q4 | 91.8% - 98.9% | First year, panel rotation effects |
| 2013-2019 | 28 quarters | 96.1% - 99.0% | Best results, stable sampling |
| 2020-2021 | 8 quarters | 93.0% - 97.5% | Pandemic sample changes |
| 2022-2024 | 12 quarters | 91.3% - 95.6% | Post-pandemic normalization |
| 2025 | Q1-Q3 | 88.9% - 95.9% | Most recent data |

---

## Using Modular Functions

For more control, use the individual functions:

```{r modular}
# Step 1: Just identify reference months
months <- identify_reference_month(pnadc)

# Step 2: Check determination rate by quarter
months[, .(
  total = .N,
  determined = sum(!is.na(ref_month_in_quarter)),
  rate = round(mean(!is.na(ref_month_in_quarter)) * 100, 1)
), by = .(Ano, Trimestre)]

# Step 3: Validate input data
validation <- validate_pnadc(pnadc, stop_on_error = FALSE)

# Step 4: Fetch population data manually
pop_data <- fetch_monthly_population(verbose = TRUE)

# Step 5: Calibrate weights separately
calibrated <- calibrate_monthly_weights(merged_data, pop_data)
```

---

## Performance

The package is optimized for large datasets:

| Dataset Size | Rows | Time (basic) | Time (weights) |
|--------------|------|--------------|----------------|
| 1 quarter | ~570K | ~5 sec | ~8 sec |
| 1 year | ~2.3M | ~26 sec | ~34 sec |
| 14 years (2012-2025) | 28.4M | 6.3 min | 8.1 min |

Tips for best performance:

- Use `data.table` directly (automatic conversion from `data.frame` adds overhead)
- Load only required columns when reading data
- Process multiple years together (cross-quarter aggregation improves accuracy)

---

## Tips and Best Practices

1. **Process multiple quarters together**: The algorithm benefits from cross-quarter aggregation. Processing 2012-2025 together gives ~96% determination vs ~70% per-quarter.

2. **Start with reference month identification**: You don't always need monthly weights. Often just knowing the reference month is enough.

3. **Check determination rates by year**: Rates should be 96-99% for 2013-2019. Lower rates may indicate data issues.

   ```{r check-rates}
   crosswalk[, .(rate = mean(!is.na(ref_month_in_quarter))), by = .(Ano, Trimestre)]
   ```

4. **Handle indeterminate observations**: Decide whether to exclude them or use quarterly-level analysis for those cases.

5. **Use `weight_monthly` for general analysis**: The rake-weighted output is appropriate for most purposes.

---

## Further Reading

- [IBGE PNADC Documentation](https://www.ibge.gov.br/estatisticas/sociais/trabalho/9171-pesquisa-nacional-por-amostra-de-domicilios-continua-mensal.html)
- Package function reference: `?mensalizePNADC`, `?identify_reference_month`
- Source code: [GitHub repository](https://github.com/antrologos/mensalizePNADC)
